### 2023-03-01 Manuscript/Draft1.Rmd
manuscript updates

-- Used MixFishSim. Describe edits made to package

We use the R package *MixFishSim* (MFS) to model our populations [@dolder2020highly]. MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. We formulate the following inputs for the MFS package to address our research question. [[perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper]]

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area [@dolder2020highly]. Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits. [[not sure we need to highlight the recruitment model, might be able to just use the table because Bev-Holt formulation is well known in fisheries]]**BL: MFS has 2 recruitment options and I edited the package to use previous years biomass** We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics. A full list of parameters used in our model can be seen below in Tables \@ref(paramsALL) and \@ref(tab:paramsSCENARIOS).

@@

Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$). We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured]. The 2012 data was repeated yearly in the constant temperature scenario because displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values. We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 temperature data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models. **dont forget to include gif in final submission**

@@

In equation (1), $Hab^2_{J,s}$ is constant for the duration of the simulation, while $Tol_{c,s,wk}$ changes each week. Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns. Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation. Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences. [[**BL: included in above paragraph** may want to mention here or in the discussion that we used repeating 2012 to simplify the simulations and reduce the number of factors explored that would have occurred with different annual cycles]]

@@

We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time. Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius [[**BL: included in above paragraph** need some justification for this large change, wanted to be extreme because if the extreme doesn't make a difference, then more reasonable increases wouldn't make a difference]] over the duration of the 20 year simulation. We therefore simulate 6 scenarios for each population.

@@

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the Spring and Fall surveys take place and the number of the samples taken reflect true target values for each strata. Most strata contain enough cells that we were able to generate a list of sampling locations without replacement, allowing us to sample a unique random cell in each survey over the duration of the simulation. For strata that contain small numbers of cells, we generate a random sampling locations with replacement of cells previously sampled. We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
**BL: added both to above paragraph**
[[should we mention here or later that we did both with and without measurement error analyses?]]
[[need to make sure that we don't lose sight of random sampling with the following sentences]]

--Stratified mean vs VAST with and without covariates
[[could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators]] 
[[may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations]]

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock. There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity. The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates. Model-based abundance estimates are often derived from model framewroks [[frameworks]] such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects. The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model [@thorson2019guidance]. VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and [[can?]] take on the order of hours to complete. We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.

>>>>

-- Used MixFishSim.

We use the R package *MixFishSim* (MFS) to model our populations [@dolder2020highly]. MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. We formulate the following inputs for the MFS package to address our research question.

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area [@dolder2020highly]. Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits. We chose to represent recruitment in the model using a Beverton-Holt formulation **cite**. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics. A full list of parameters used in our model can be seen below in Tables \@ref(paramsALL) and \@ref(tab:paramsSCENARIOS).

@@

Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$). We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured]. We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models. **dont forget to include gif in final submission**

@@

In equation (1), $Hab^2_{J,s}$ is constant for the duration of the simulation, while $Tol_{c,s,wk}$ changes each week. Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns. Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation. Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.

@@

We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time. Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation. We therefore simulate 6 scenarios for each population.

@@

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the Spring and Fall surveys take place, and the number of the samples taken reflect true target values for each strata. Most strata contain enough cells that we were able to generate a list of sampling locations without replacement, allowing us to sample a unique random cell in each survey over the duration of the simulation. For strata that contain small numbers of cells, we generate a random sampling locations with replacement of cells previously sampled. We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.

--Stratified mean vs VAST with and without covariates

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each fish stock. There are a number of approaches to obtain abundance estimates including traditional design-based estimates to model-based estimates that range in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples or allow the inclusion of environmental influences. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.

We compare abundance estimates from stratified mean calculations to those derived from the Vector Autoregressive Spatio-Temporal (VAST) model. The stratified mean is a design-based approach that calculates the geometric mean catch per tow and has traditionally been used with stratified random sample designs. VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) [@thorson2019guidance]. If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete. We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.

### 2023-03-01 Manuscript/Draft1.tex
manuscript updates

Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration's (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts. Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (Azarovitz 1981; Politis et al. 2014). The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits

The survey takes place twice each year- once in the spring and again in the fall. Since most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling during approximately the same 2-3 week time period in each season. {[}{[}the survey for the entire region takes \textasciitilde8 weeks to complete each season, the GB strata get covered in a similar time period each season because the survey starts in the south and moves north. But there are times when logistics, e.g., storms, Covid, cause a change in the timing of any particular stratum being sampled. This timing issue has come up in the past and we've looked at it using the dates of sampled stations within a particular stock definition. See for example GBYT\_survey\_timing\_cdf.png taken from the 2018 assessment of Georges Bank yellowtail flounder that I copied into a newly created ``miscellaneous'' folder on the Google Drive. Given all this, I suggest modifying the wording to something like ``NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season''{]}{]}

@@

-- Used MixFishSim. Describe edits made to package

We use the R package \emph{MixFishSim} (MFS) to model our populations (Dolder et al. 2020). MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. We formulate the following inputs for the MFS package to address our research question. {[}{[}perhaps for discussion, but we should mention somewhere that MFS only tracks biomass not age structure, the latter is important for stock assessments but not considered in this paper{]}{]}

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area (Dolder et al. 2020). Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits. {[}{[}not sure we need to highlight the recruitment model, might be able to just use the table because Bev-Holt formulation is well known in fisheries{]}{]}\textbf{BL: MFS has 2 recruitment options and I edited the package to use previous years biomass} We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics. A full list of parameters used in our model can be seen below in Tables \ref{paramsALL} and \ref{tab:paramsSCENARIOS}.

@@

Each species is assumed to have normally distributed temperature preferences (\(N(\mu,\sigma)\)). We assume Yellowtail Flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\). We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al. 2006). The 2012 data was repeated yearly in the constant temperature scenario because displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values. We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 temperature data was also transform to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models. \textbf{dont forget to include gif in final submission}

@@

In equation (1), \(Hab^2_{J,s}\) is constant for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week. Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns. Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation. Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences. {[}{[}\textbf{BL: included in above paragraph} may want to mention here or in the discussion that we used repeating 2012 to simplify the simulations and reduce the number of factors explored that would have occurred with different annual cycles{]}{]}

@@

We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time. Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius {[}{[}\textbf{BL: included in above paragraph} need some justification for this large change, wanted to be extreme because if the extreme doesn't make a difference, then more reasonable increases wouldn't make a difference{]}{]} over the duration of the 20 year simulation. We therefore simulate 6 scenarios for each population.

@@

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the Spring and Fall surveys take place and the number of the samples taken reflect true target values for each strata. Most strata contain enough cells that we were able to generate a list of sampling locations without replacement, allowing us to sample a unique random cell in each survey over the duration of the simulation. For strata that contain small numbers of cells, we generate a random sampling locations with replacement of cells previously sampled. We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.
\textbf{BL: added both to above paragraph}
{[}{[}should we mention here or later that we did both with and without measurement error analyses?{]}{]}
{[}{[}need to make sure that we don't lose sight of random sampling with the following sentences{]}{]}

--Stratified mean vs VAST with and without covariates
{[}{[}could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators{]}{]}
{[}{[}may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations{]}{]}

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock. There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity. The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates. Model-based abundance estimates are often derived from model framewroks {[}{[}frameworks{]}{]} such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects. The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model (Thorson 2019). VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and {[}{[}can?{]}{]} take on the order of hours to complete. We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.

>>>>

Fish stocks in this highly productive and economically important region are managed by the National Oceanic and Atmospheric Administration's (NOAA) Northeast Fisheries Science Center (NEFSC) in Woods Hole, Massachusetts. Federal scientists assess the health and abundance of each commercial fish stock using fishery-independent bottom trawl survey data that has been collected by NOAA throughout the region since 1963 (Azarovitz 1981; Politis et al. 2014). The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits. The survey takes place twice each year- once in the spring and again in the fall. Since most spatial analyses and projections of future distributions typically assume a constant survey catchability and/or availability over time, NOAA's survey design includes sampling Georges Bank during approximately the same 3-4 week time period in each season.

@@

-- Used MixFishSim.

We use the R package \emph{MixFishSim} (MFS) to model our populations (Dolder et al. 2020). MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. We formulate the following inputs for the MFS package to address our research question.

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area (Dolder et al. 2020). Individual terms in the formulation account for growth of mature adults, natural and fishing mortality, and the addition of new recruits. We chose to represent recruitment in the model using a Beverton-Holt formulation \textbf{cite}. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics. A full list of parameters used in our model can be seen below in Tables \ref{paramsALL} and \ref{tab:paramsSCENARIOS}.

@@

Each species is assumed to have normally distributed temperature preferences (\(N(\mu,\sigma)\)). We assume Yellowtail Flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\). We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al. 2006). We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models. \textbf{dont forget to include gif in final submission}

@@

In equation (1), \(Hab^2_{J,s}\) is constant for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week. Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns. Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation. Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.

@@

We consider 20 year simulations under three population parameter scenarios for each of our three species- a scenario where parameters result in each population increasing over time, one where the populations are relatively constant over time, and a scenario where the parameter combination results in each population decreasing over time. Each of these three scenarios is paired with a temperature gradient that repeats as well as one that increasing roughly 5 degrees Celsius over the duration of the 20 year simulation. We therefore simulate 6 scenarios for each population.

@@

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the Spring and Fall surveys take place, and the number of the samples taken reflect true target values for each strata. Most strata contain enough cells that we were able to generate a list of sampling locations without replacement, allowing us to sample a unique random cell in each survey over the duration of the simulation. For strata that contain small numbers of cells, we generate a random sampling locations with replacement of cells previously sampled. We then use the biomass collected from our samples in contemporary population indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compare the error calculated from each estimation method.

--Stratified mean vs VAST with and without covariates

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each fish stock. There are a number of approaches to obtain abundance estimates including traditional design-based estimates to model-based estimates that range in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples or allow the inclusion of environmental influences. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.

We compare abundance estimates from stratified mean calculations to those derived from the Vector Autoregressive Spatio-Temporal (VAST) model. The stratified mean is a design-based approach that calculates the geometric mean catch per tow and has traditionally been used with stratified random sample designs. VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete. We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
